Cover Times of Random Walks

### 2019-04-30 paper.tex
dots and indices

We write the cover time in terms of $X_i$ and use linearity of expectation.
\begin{align}
\tcov&= \E[X_1+X_2+X_3+\ldots+X_{n+1}] = \sum_{i=1}^n \E[X_i] \nonumber
\end{align}

@@

\begin{comment}
It follows that $X_i$ is geometric with distribution
\begin{align}
\pr(X_i = k) = (1-p_i)^{k-1}p_i, \nonumber
\end{align}
where $p_i$ is the probability of moving to a previously unvisited vertex.
\end{comment}
We now find the probability $p_i$ that 
we go to a previously unvisited vertex is $(n-i)/(n-1)$.
This is because there are $n-i$ unvisited vertices and a total of $n-1$ adjacent vertices.
It follows that $\E[X_i] = (ln-1)(n-i)$.
Then
\begin{align}
\tcov&=   \sum_{i=1}^n \frac{n-1}{n-i} \nonumber \\
&=  (n-1) \left(\frac{1}{n-1}+\frac{1}{n-2}+\ldots+1\right). \nonumber
\end{align}

A natural interpretation of the cover time on a complete graph is 
the so-called coupon collecting problem.
There are $r$ different coupons that are randomly packaged in cereal boxes.
An avid fan buys a box of cereal every day.
We want to know how many days until the fan has collected each coupon.
The strategy is to think of every coupon as a vertex on the complete graph.
We then move from vertex to vertex with uniform probability.
The difference between the coupon collecting problem and the complete graph
is that we have self-edges
whereas before we had to leave our current vertex at each step.
Now, we can stay in the same vertex (provided our collector found the same coupon
two days in a row).
We apply our approach to the complete graph and substitute the $r$ possible coupons
we can find for the $n-1$ vertices we could move to.
Then
\begin{align}
\tcov&=  r \left(\frac{1}{r}+\frac{1}{r-1}+\ldots+1\right). \nonumber
\end{align}

@@

\begin{align}
\tcov&= 2(n-1) \left(\frac{1}{n-1}+\frac{1}{n-2}+\ldots+1\right). \nonumber
\end{align}

@@

\begin{align}
(e_i - e_1) &= (i-1) e_1 - 2[1+2+\ldots+i-1] \nonumber \\
e_i &= i(N-1) - 2 \frac{(i-1)i}{2} \nonumber\\
&= i(N-1 - i + 1) = i(N-i)
\label{eq:line}
\end{align}

@@

We write
\begin{align}
\E_0[T_N] &= \E_0[T_1] + \E_1[T_2] + \ldots + \E_{i}[T_{i+1}] + \ldots + \E_{N-1}[T_N]. \nonumber
\end{align}

@@

Then
\begin{align}
\E_0[T_N] &= 1 + 3 + \ldots + 2i - 1 + \ldots + 2(N-1) -1 \nonumber \\
&= N^2 \nonumber
\end{align}

@@

Let $\{a, a+1, \dots, b-1, b\}$ 
denote the set of vertices we have visited.
Since the walk is on a line, we can $[a,b]$ for the range
of the walk where $a$ is the smallest 
and $b$ is the largest vertex we have visited.
Notice that if $a \leq -n$ or $b \geq n$, we have covered all $n$ vertices
on the simple cycle and conclude the random walk.

In a slight abuse of notation, let $T_k$ be the 
number of steps until we reach $k$ distinct vertices.
Formally, $T_k = \min\{t: b-a+1 = k\}$.
Then by telescoping,
\begin{align}
\tcov &= \E[T_n] = \E[(T_2 - T_1) + (T_3 - T_2) + \dots + 
(T_k - T_{k-1}) + \dots + (T_n - T_{n-1})] \nonumber
\end{align}

@@

By linearity of expectation,
\begin{align}
\tcov &=  (\E[T_2] - \E[T_1]) + \dots +
(\E[T_{k+1}] - \E[T_{k}]) + \dots + (\E[T_n] - \E[T_{n-1}]) \nonumber\\
&= 1 + 2 + \dots + k-1 + \dots + n-1  = \frac{1}{2}n(n-1). \nonumber
\end{align}

@@

\begin{theorem}[Matthews Upper Bound]
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews_up}
\end{theorem}
\begin{align}
\tcov &\leq \hit \left(1 + \frac{1}{2} + \frac{1}{3} + \ldots + \frac{1}{n-1} \right). \nonumber
\end{align}

\begin{proof}
Without loss of generality, label the vertices $\{1, 2, \ldots, n\}$.
We may assume that the walk started at vertex $n$.
Let $\sigma$ be a uniform permutation of the unvisited $n-1$ vertices.
The strategy is to look for the vertices in order $\sigma$.

@@

We begin by writing the cover time in a clever way
(similar to the strategy we employed on the cycle graph):
\begin{align}
\tcov &= \E_n[{T_{n-1}}]  \nonumber \\
&= \E_n[T_1 + (T_2 - T_1) + \ldots + 
(T_k - T_{k-1}) + \ldots + (T_{n-1} - T_{n-2})] . \nonumber
\end{align}

@@

Now consider the $k^{th}$ term of the sum for $1 < k \leq n-1$.
By conditional expectation,

\begin{align}
\E_n[T_k - T_{k-1}] &=
\sum_{i=1}^k \E_n[T_k - T_{k-1} | L(k) = \sigma_i]
P(L_k = \sigma_i). \nonumber 
\end{align}

@@

Putting it all together,
\begin{align}
\tcov \leq \hit \left( 1+ \frac{1}{2} + \ldots + \frac{1}{k} 
+ \ldots + \frac{1}{n-1} \right). \nonumber
\end{align}
\end{proof}

@@

\begin{theorem}[Matthews Lower Bound]
Let $A \subseteq V(G)$. Define $\tmina = \min_{a,b\in A, a \neq b} \E_b[T_a]$.
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews_low}
\end{theorem}
\begin{align}
\tcov &\geq \tmina \left(1 + \frac{1}{2} + \frac{1}{3} + \ldots +
\frac{1}{|A| -1} \right). \nonumber
\end{align}

@@

Let $T_k$ be the first time that vertices $\sigma_1, \sigma_2, \ldots, \sigma_k$
have all been visited.
For ease of notation, set $L_k = X_{T_k}$.

We again write the cover time in a clever way:
\begin{align}
\tcov &\geq \E_x[T_1 + (T_2 - T_1) + \ldots + 
(T_k - T_{k-1}) + \ldots + (T_{|A|} - T_{|A|-1})] . \nonumber
\end{align}

@@

Now consider the $k^{th}$ term of the sum for $k > 1$.
By conditional expectation,
\begin{align}
\E_x[T_k - T_{k-1}] &=
\sum_{i=1}^k \E_x[T_k - T_{k-1} | L(k) = \sigma_i]
P(L_k = \sigma_i). \nonumber 
\end{align}

@@

Putting it all together,
\begin{align}
\tcov &\geq \tmina \left( 1 - \frac{1}{|A|}+ \frac{1}{2} + \ldots + \frac{1}{k} 
+ \ldots + \frac{1}{|A|-1} + \frac{1}{|A|}\right). \nonumber \\
&= \tmina \left( 1 + \frac{1}{2} + \ldots + \frac{1}{k} 
+ \ldots + \frac{1}{|A|-1} \right). \nonumber
\end{align}
\end{proof}

@@

\cite{LP17}
Consider an unweighted binary tree as in \cref{fig:binary} with a single root $\rho$.
A vertex has two children if it is distance less than $k$
from the root and no children if it is a leaf (distance $k$ from the root).
There are $n = 1 + 2 + 4 + \ldots + 2^k = 2^{k+1} - 1$ vertices.

@@

and \cref{thm:matthews_up} gives
\begin{align}
\tcov &\leq 2(n-1)k\left(1 + \frac{1}{2} + \ldots + \frac{1}{n} \right) \nonumber \\
&\approx (2 - \frac{2}{n}) \cdot n k \log n \approx (2 - \frac{2}{n}) n k \log 2^{k+1} -1 \nonumber \\
&\approx (2  + o(1)) \cdot n k^2 \log 2 \nonumber
\end{align}

@@

In the worst case, $h' = h-1$.
Then by \cref{thm:matthews_low},
\begin{align}
\tcov &\geq 2 (n-1)(k-h+1)\left(1 + \frac{1}{2} + \ldots + \frac{1}{2^h -1} \right) \nonumber \\
&\approx (2 + o(1)) \cdot n (k-h) h \log 2 \nonumber.
\end{align}

@@

By the principle of inclusion-exclusion,
\begin{align}
p_0(r,n) &= {n \choose 0} 1^r - {n \choose 1} (1-1/n)^r + \ldots \nonumber\\
&= \sum_{k=0}^n (-1)^k {n \choose k} (1-k/n)^r. \label{eqn:p0}
\end{align}

@@

As $n \rightarrow \infty$, $\left(1-\frac{m}{n}\right)^m \rightarrow 1$
and $1/m!$ is constant.
We use the Taylor series of $\log(1-t)$ for $0 \leq t \leq 1/2$
to bound the rest.
Notice that
\begin{align}
\log(1-t) &= 0 - t - t^2/2 -t^3/3 - \ldots \nonumber\\
&\geq -t -t^2/2(1 + 1/2 + 1/4 + \ldots) = -t - t^2 \nonumber
\end{align}
and so we have
\begin{align}
\log \left(n^m\left(1-\frac{m}{n}\right)^r \right)
&= m \log n + r \log\left(1-\frac{m}{n}\right) \nonumber \\
& \geq m \log n - r\frac{m}{n} - r\frac{m}{n^2}. \nonumber
\end{align}

>>>>

We write the cover time in terms of $X_i$ and use linearity of expectation.
\begin{align}
\tcov&= \E[X_1+X_2+X_3+\cdots+X_{n+1}] = \sum_{i=1}^n \E[X_i] \nonumber
\end{align}

@@

\begin{comment}
It follows that $X_i$ is geometric with distribution
\begin{align}
\pr(X_i = k) = (1-p_i)^{k-1}p_i, \nonumber
\end{align}
where $p_i$ is the probability of moving to a previously unvisited vertex.
\end{comment}
We now find the probability $p_i$ that 
we go to a previously unvisited vertex is $(n-i)/(n-1)$.
This is because there are $n-i$ unvisited vertices and a total of $n-1$ adjacent vertices.
It follows that $\E[X_i] = (n-1)(n-i)$.
Then
\begin{align}
\tcov&=   \sum_{i=1}^{n-1} \frac{n-1}{n-i} \nonumber \\
&=  (n-1) \left(\frac{1}{n-1}+\frac{1}{n-2}+\cdots+1\right). \nonumber
\end{align}

A natural interpretation of the cover time on a complete graph is 
the so-called coupon collecting problem.
There are $r$ different coupons that are randomly packaged in cereal boxes.
An avid fan buys a box of cereal every day.
We want to know how many days until the fan has collected each coupon.
The strategy is to think of every coupon as a vertex on the complete graph.
We then move from vertex to vertex with uniform probability.
The difference between the coupon collecting problem and the complete graph
is that we have self-edges
whereas before we had to leave our current vertex at each step.
Now, we can stay in the same vertex (provided our collector found the same coupon
two days in a row).
We apply our approach to the complete graph and substitute the $r$ possible coupons
we can find for the $n-1$ vertices we could move to.
Then
\begin{align}
\tcov&=  r \left(\frac{1}{r}+\frac{1}{r-1}+\cdots+1\right). \nonumber
\end{align}

@@

\begin{align}
\tcov&= 2(n-1) \left(\frac{1}{n-1}+\frac{1}{n-2}+\cdots+1\right). \nonumber
\end{align}

@@

\begin{align}
(e_i - e_1) &= (i-1) e_1 - 2[1+2+\cdots+i-1] \nonumber \\
e_i &= i(N-1) - 2 \frac{(i-1)i}{2} \nonumber\\
&= i(N-1 - i + 1) = i(N-i)
\label{eq:line}
\end{align}

@@

We write
\begin{align}
\E_0[T_N] &= \E_0[T_1] + \E_1[T_2] + \cdots + \E_{i}[T_{i+1}] + \cdots + \E_{N-1}[T_N]. \nonumber
\end{align}

@@

Then
\begin{align}
\E_0[T_N] &= 1 + 3 + \cdots + 2i - 1 + \cdots + 2(N-1) -1 \nonumber \\
&= N^2 \nonumber
\end{align}

@@

Let $\{a, a+1, \ldots, b-1, b\}$ 
denote the set of vertices we have visited.
Since the walk is on a line, we can $[a,b]$ for the range
of the walk where $a$ is the smallest 
and $b$ is the largest vertex we have visited.
Notice that if $a \leq -n$ or $b \geq n$, we have covered all $n$ vertices
on the simple cycle and conclude the random walk.

In a slight abuse of notation, let $T_k$ be the 
number of steps until we reach $k$ distinct vertices.
Formally, $T_k = \min\{t: b-a+1 = k\}$.
Then by telescoping,
\begin{align}
\tcov &= \E[T_n] = \E[(T_2 - T_1) + (T_3 - T_2) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_n - T_{n-1})] \nonumber
\end{align}

@@

By linearity of expectation,
\begin{align}
\tcov &=  (\E[T_2] - \E[T_1]) + \cdots +
(\E[T_{k+1}] - \E[T_{k}]) + \cdots + (\E[T_n] - \E[T_{n-1}]) \nonumber\\
&= 1 + 2 + \cdots + k-1 + \cdots + n-1  = \frac{1}{2}n(n-1). \nonumber
\end{align}

@@

\begin{theorem}[Matthews Upper Bound]
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews_up}
\end{theorem}
\begin{align}
\tcov &\leq \hit \left(1 + \frac{1}{2} + \frac{1}{3} + \cdots + \frac{1}{n-1} \right). \nonumber
\end{align}

\begin{proof}
Without loss of generality, label the vertices $\{1, 2, \dots, n\}$.
We may assume that the walk started at vertex $n$.
Let $\sigma$ be a uniform permutation of the unvisited $n-1$ vertices.
The strategy is to look for the vertices in order $\sigma$.

@@

We begin by writing the cover time in a clever way
(similar to the strategy we employed on the cycle graph):
\begin{align}
\tcov &= \E_n[{T_{n-1}}]  \nonumber \\
&= \E_n[T_1 + (T_2 - T_1) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_{n-1} - T_{n-2})] . \nonumber
\end{align}

@@

Now consider the $k^{th}$ term of the sum for $1 < k \leq n-1$.
By total expectation,
\begin{align}
\E_n[T_k - T_{k-1}] &=
\sum_{i=1}^k \E_n[T_k - T_{k-1} | L(k) = \sigma_i]
P(L_k = \sigma_i). \nonumber 
\end{align}

@@

Putting it all together,
\begin{align}
\tcov \leq \hit \left( 1+ \frac{1}{2} + \cdots + \frac{1}{k} 
+ \cdots + \frac{1}{n-1} \right). \nonumber
\end{align}
\end{proof}

@@

\begin{theorem}[Matthews Lower Bound]
Let $A \subseteq V(G)$. Define $\tmina = \min_{a,b\in A, a \neq b} \E_b[T_a]$.
Let $(X_t)$ be a random walk on a graph with $n$ vertices. Then \label{thm:matthews_low}
\end{theorem}
\begin{align}
\tcov &\geq \tmina \left(1 + \frac{1}{2} + \frac{1}{3} + \cdots +
\frac{1}{|A| -1} \right). \nonumber
\end{align}

@@

Let $T_k$ be the first time that vertices $\sigma_1, \sigma_2, \cdots, \sigma_k$
have all been visited.
For ease of notation, set $L_k = X_{T_k}$.

We again write the cover time in a clever way:
\begin{align}
\tcov &\geq \E_x[T_1 + (T_2 - T_1) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_{|A|} - T_{|A|-1})] . \nonumber
\end{align}

@@

Now consider the $k^{th}$ term of the sum for $k > 1$.
By total expectation,
\begin{align}
\E_x[T_k - T_{k-1}] &=
\sum_{i=1}^k \E_x[T_k - T_{k-1} | L(k) = \sigma_i]
P(L_k = \sigma_i). \nonumber 
\end{align}

@@

Putting it all together,
\begin{align}
\tcov &\geq \tmina \left( 1 - \frac{1}{|A|}+ \frac{1}{2} + \cdots + \frac{1}{k} 
+ \cdots + \frac{1}{|A|-1} + \frac{1}{|A|}\right). \nonumber \\
&= \tmina \left( 1 + \frac{1}{2} + \cdots + \frac{1}{k} 
+ \cdots + \frac{1}{|A|-1} \right). \nonumber
\end{align}
\end{proof}

@@

\cite{LP17}
Consider an unweighted binary tree as in \cref{fig:binary} with a single root $\rho$.
A vertex has two children if it is distance less than $k$
from the root and no children if it is a leaf (distance $k$ from the root).
There are $n = 1 + 2 + 4 + \cdots + 2^k = 2^{k+1} - 1$ vertices.

@@

and \cref{thm:matthews_up} gives
\begin{align}
\tcov &\leq 2(n-1)k\left(1 + \frac{1}{2} + \cdots + \frac{1}{n} \right) \nonumber \\
&\approx (2 - \frac{2}{n}) \cdot n k \log n \approx (2 - \frac{2}{n}) n k \log 2^{k+1} -1 \nonumber \\
&\approx (2  + o(1)) \cdot n k^2 \log 2 \nonumber
\end{align}

@@

In the worst case, $h' = h-1$.
Then by \cref{thm:matthews_low},
\begin{align}
\tcov &\geq 2 (n-1)(k-h+1)\left(1 + \frac{1}{2} + \cdots + \frac{1}{2^h -1} \right) \nonumber \\
&\approx (2 + o(1)) \cdot n (k-h) h \log 2 \nonumber.
\end{align}

@@

By the principle of inclusion-exclusion,
\begin{align}
p_0(r,n) &= {n \choose 0} 1^r - {n \choose 1} (1-1/n)^r + \cdots \nonumber\\
&= \sum_{k=0}^n (-1)^k {n \choose k} (1-k/n)^r. \label{eqn:p0}
\end{align}

@@

As $n \rightarrow \infty$, $\left(1-\frac{m}{n}\right)^m \rightarrow 1$
and $1/m!$ is constant.
We use the Taylor series of $\log(1-t)$ for $0 \leq t \leq 1/2$
to bound the rest.
Notice that
\begin{align}
\log(1-t) &= 0 - t - t^2/2 -t^3/3 - \cdots \nonumber\\
&\geq -t -t^2/2(1 + 1/2 + 1/4 + \cdots) = -t - t^2 \nonumber
\end{align}
and so we have
\begin{align}
\log \left(n^m\left(1-\frac{m}{n}\right)^r \right)
&= m \log n + r \log\left(1-\frac{m}{n}\right) \nonumber \\
& \geq m \log n - r\frac{m}{n} - r\frac{m}{n^2}. \nonumber
\end{align}
